Beyond May: Complexity-stability relationships \\ in disordered dynamical systems

### 2023-05-05 papers/beyond-may.tex
add link to repo
update draft

\begin{acknowledgments}
We thank Ian Hatton, Matthieu Barbier and Ada Altieri for productive discussions on ecological power laws and the diversity-stability debate. 
\end{acknowledgments}

>>>>

\begin{acknowledgments}
We thank Ian Hatton, Matthieu Barbier and Ada Altieri for productive discussions on ecological power laws and the diversity-stability debate and Vadim Karatayev and Daniel Reuman for feedbacks on the manuscript.
O.M. was partly supported by NSF BIO OCE grant 2023473.
\end{acknowledgments}

### 2023-05-05 papers/conclusion.tex
update draft

The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and in other fields. 
In this letter, we have showed that the condition $\sigma\sqrt{N}< \mu_s - \mu$ (popularized by May's slogan ``complexity begets instability") does not provide a complete picture of the relationship between complexity and stability. 
In particular, we have seen that generalized Lotka-Volterra model, often cited in support of May's prediction, corresponds to a special cancellation in the more general stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
In models where the self-dynamics is more strongly regulated than the cross-interactions ($\alpha < \beta$, which might be more appropriate for ecological modeling \cite{Hatton2023}), the opposite behavior is observed: stability becomes \emph{more} likely with increasing diversity $N$.

>>>>

\emph{Discussion.---}The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and in other fields. 
In this letter, we have showed that the condition $\sigma\sqrt{N}< \mu_s - \mu$ (popularized by May's slogan ``complexity begets instability") does not provide a complete picture of the relationship between complexity and stability. 
In particular, we have seen that generalized Lotka-Volterra model, often cited in support of May's prediction, corresponds to a special cancellation in the more general stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
In models where the self-dynamics is more strongly regulated than the cross-interactions, the opposite behavior is observed: stability becomes \emph{more} likely with increasing diversity $N$.

GLV is special also for another reason: for homogenous interaction, dependence on the number of degrees of freedom is lost in the stability condition.
We showed that in general this dependence is present, and its effect on the complexity-stablity property is preserved for heterogeneous interaction. 
The simple condition $\psi(x^*)g(x^*)h'(x^*)>0$ can therefore be employed to gain informations on systems desceribed by Eq.~\eqref{dynamics}.
Future work can be done in the direction of extending the equations for testing complexity-stability for cases in which we can not decompose the two-body interactions, or to include higher order interaction, generalizing, e.g., the analysis of recent results~\cite{Gibbs2022} based on GLV.

Our general conclusion that for large random dynamical systems, an increase in dimensionality can lead to either stabilization or destabilization, without invoking specific system topology, bares significant implications for the study and control of complex systems.

### 2023-05-05 papers/equilibrium.tex
update draft

\emph{Random interactions: equilibrium distribution.---}Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=-x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
(Any coefficients can be reabsorbed in the statistics of $A$ and by a rescaling of time. 
We factor out a minus sign in order to connect more directly with competition models.)

\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10-diversity.pdf}
    \caption{For a system in the ``anti-May" phase, an increase in $N$,
    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane, will be stabilizing rather than destabilizing.
    Here we show results from the numerical resolution of the dynamical system~\eqref{dynamics}. Stability is defined as full stable coexistence. The red line is computed obtained using DMFT and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
    Parameters values are $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=50$ and $10$ replicates for the simulations.}
    \label{fig: stability line + sims}
\end{figure}

Following e.g. Ref.~\cite{Roy2019}, we can derive from the evolution equation~\eqref{dynamics} a dynamical mean field theory (DMFT) 
which describes the ensemble of $N\gg 1$ variables
by means of a single, self-consistent representative stochastic trajectory
\begin{equation}
    \dot{x} = x^{\alpha}-A_sx^{\beta+\gamma}-x^{\beta}\big( \mu N \langle x^{\gamma}\rangle + \sigma \sqrt{N} \eta\big) \, ,
\label{eq: dmft}
\end{equation}
where $A_s$ is a random variable with the statistics of $A_{ii}$ and $\eta$ is a Gaussian noise with zero mean and correlation $\langle x^{\gamma}(t)x^{\gamma}(s)\rangle$.

\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{figs/alpha-beta.pdf}
    \caption{The ``May" and ``anti-May" phases described in the
    case of uniform interactions are robust with respect to random interactions.
    Here we simulated system~\eqref{dynamics} in the power law case for a fixed value of
    $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates 
    and evaluated $\sigma_c$, defined here as
    the value above which full stable coexistence is less probable than 50\%,
    at varying $\alpha$ and $\beta$.
    No amount of heterogeneity is allowed in the lower triangle because we have $\mu=\mu_s$, while an increasing value of $\sigma_c$
    is found as $\beta$ becomes bigger than $\alpha$.
    We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which instability is triggered also in the upper triangle,
    does not imply that increasing $N$ will destabilize the system, because $\mu N$ would also change.
    }
    \label{fig: alpha-beta}
\end{figure}

At equilibrium, we can write
\begin{equation}
    x_*^{\alpha-\beta} - A_s x_*^{\gamma}= \big( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N\langle x_*^{2\gamma}\rangle}\xi\big) \, ,
\end{equation} 
where $\xi$ is now a standard normal random variable.
The equation above can be solved for $x_*$ for specific values of
$(\alpha-\beta)/\gamma$, or if $A_s=0$.
(In the following we show that neglecting $A_s$ even when it is not zero does not affect noticeably the quality of the approximation.) In this case the stationary solution is given by 
\begin{equation} \label{eq: cavity solution}
    x_* = \left( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N\langle x_*^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, .
\end{equation}
The equilibrium distribution $P(x_*)$ can then be obtained as the pushforward of the distribution of $\xi$: 
\begin{equation}\label{eq: dist general}
    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\sigma^2 N\langle x_*^{2\gamma}\rangle}}
    \exp{\left\{-\frac{(x_*^{\alpha-\beta}-\mu N\langle x_*^{\gamma}\rangle)^2}{2\sigma^2N\langle x_*^{2\gamma}\rangle}\right\}} \, ,
\end{equation}
where the expectation values must be computed self-consistently.

@@

Equipped with the equilibrium distribution $P(x^*)$, we can compute the maximal heterogeneity 
$\sigma_c$, for a given $N$ and $\mu$, for which the linear stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$ holds. In Fig.~\ref{fig: stability line + sims} the result in portrayed in the $(\sigma \sqrt{N},\mu N)$ plane,
compared with results from simulations (where stability is defined
as stable, full, coexistence of the system).
Everything else fixed, an increase in the number of
degrees of freedom $N$ will move the system along a square-root
trajectory in that plane, eventually entering the stable region where $\sigma_c > \sigma$. 
In other words, we are in the
``anti-May" phase.
(For comparison the stability boundary for GLV is a horizontal line in the $(\sigma \sqrt{N},\mu N)$ plane, defined by $\sigma_c\sqrt{N} = \mu_s$ \cite{bunin2017ecological}.)

Figure~\ref{fig: alpha-beta} shows results of simulations 
for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, 
at fixed $\mu N$ and $\gamma = 1$, 
defined as the value above which
full stable coexistence is less probable than $50\%$.
(In these simulations we use $\mu_s = \mu$, which would never be stable in the GLV model.) This plot illustrates the transition between a ``May" phase for $\alpha \geq \beta$ (where stability decreases with $N$, $\mu$, and $\sigma$) and an ``anti-May" phase for $\alpha < \beta$, even in the presence of heterogeneous interactions.

>>>>

\emph{Random interactions: equilibrium distribution.---}Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
(Any coefficients can be reabsorbed in the statistics of $A$ and by a rescaling of time.)
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10-diversity-increase.pdf}
    \caption{For a system in the ``anti-May" phase, an increase in $N$,
    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane (green dashed line), will be stabilizing rather than destabilizing.
    Here we show results from the numerical resolution of the dynamical system~\eqref{dynamics}. Stability is defined as full stable coexistence. The red line is computed obtained using DMFT and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
    Parameters values are $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=50$ and $10$ replicates for the simulations. For the line representing the increase in $N$ we fixed $\mu=0.1$ and $\sigma=0.75$.}
    \label{fig: stability line + sims}
\end{figure}
Following e.g. Ref.~\cite{Roy2019}, we can derive from the evolution equation~\eqref{dynamics} a dynamical mean field theory (DMFT) 
which describes the ensemble of $N\gg 1$ variables
by means of a single, self-consistent representative stochastic trajectory
\begin{equation}
    \dot{x} = x^{\alpha}-A_sx^{\beta+\gamma}-x^{\beta}\big( \mu N \langle x^{\gamma}\rangle + \sigma \sqrt{N} \eta\big) \, ,
\label{eq: dmft}
\end{equation}
where $A_s$ is a random variable with the statistics of $A_{ii}$ and $\eta$ is a Gaussian noise with zero mean and correlation $\langle x^{\gamma}(t)x^{\gamma}(s)\rangle$. 
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{figs/alpha-beta.pdf}
    \caption{The ``May" and ``anti-May" phases described in the
    case of uniform interactions are robust with respect to random interactions.
    Here we simulated system~\eqref{dynamics} in the power law case for a fixed value of
    $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates 
    and evaluated $\sigma_c$, defined here as
    the value above which full stable coexistence is less probable than 50\%,
    at varying $\alpha$ and $\beta$.
    No amount of heterogeneity is allowed in the lower triangle because we have $\mu=\mu_s$, while an increasing value of $\sigma_c$
    is found as $\beta$ becomes bigger than $\alpha$.
    We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which instability is triggered also in the upper triangle,
    does not imply that increasing $N$ will destabilize the system, because $\mu N$ would also change.
    }
    \label{fig: alpha-beta}
\end{figure}
At equilibrium, we can write
\begin{equation}
    x_*^{\alpha-\beta} - A_s x_*^{\gamma}= \big( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N\langle x_*^{2\gamma}\rangle}\xi\big) \, ,
\end{equation} 
where $\xi$ is now a standard normal random variable.
The equation above can be solved for $x_*$ for specific values of
$(\alpha-\beta)/\gamma$, or if $A_s=0$.
(In the following we show that neglecting $A_s$ even when it is not zero does not affect noticeably the quality of the approximation.) In this case the stationary solution is given by 
\begin{equation} \label{eq: cavity solution}
    x_* = \left( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N\langle x_*^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, .
\end{equation}
The equilibrium distribution $P(x_*)$ can then be obtained as the pushforward of the distribution of $\xi$: 
\begin{equation}\label{eq: dist general}
    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\sigma^2 N\langle x_*^{2\gamma}\rangle}}
    \exp{\left\{-\frac{(x_*^{\alpha-\beta}-\mu N\langle x_*^{\gamma}\rangle)^2}{2\sigma^2N\langle x_*^{2\gamma}\rangle}\right\}} \, ,
\end{equation}
where the expectation values must be computed self-consistently.

@@

Equipped with the equilibrium distribution $P(x^*)$, we can compute the maximal heterogeneity $\sigma_c$, for a given $N$ and $\mu$, for which the linear stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$ holds. 
A convenient way to test the complexity-stability properties of the system, consists in studying how this condition compart the $(\sigma \sqrt{N},\mu N)$ plan, conventionally used to summarize the properties of large random dynamical systems~\cite{bunin2017ecological}, in stable and unstable parameter regions. 
Everything else fixed, an increase in the number of degrees of freedom $N$ will move the system along a square-root trajectory in that plane and we can test if this has stabilizing effects on the system or if the opposite is true.
In Fig.~\ref{fig: stability line + sims} the linear stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$ is portrayed in the $(\sigma \sqrt{N},\mu N)$ plane, compared with results from simulations (where stability is defined as stable, full, coexistence of the system).
An increase in the number of degrees of freedom $N$ will eventually bring the system in the stable region, or preserve stability if already stable. 
In other words, we are in the ``anti-May" phase.
(For comparison the stability boundary for GLV is a horizontal line in the $(\sigma \sqrt{N},\mu N)$ plane, defined by $\sigma_c\sqrt{N} = \mu_s$~\cite{bunin2017ecological}.)

We conclude by testing the robustness of our findings in the case of heterogeneous interactions in the paradigmatic case of power laws. Fig.~\ref{fig: alpha-beta} shows results of simulations for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, at fixed $\mu N$ and $\gamma = 1$, defined as the value above which full stable coexistence is less probable than $50\%$.
(In these simulations we use $\mu_s = \mu$, which would never be stable in the GLV model.) This plot illustrates the transition between a ``May" phase for $\alpha \geq \beta$ (where stability decreases with $N$, $\mu$, and $\sigma$) and an ``anti-May" phase for $\alpha < \beta$, even in the presence of heterogeneous interactions.

### 2023-05-05 papers/homogeneous.tex
update draft

\emph{Homogeneous interactions.---}Under what condition does~\eqref{dynamics} admit a (linearly) stable equilibrium? 
We begin with the simple case where all self-interactions have the same strength $A_{ii} = \mu_s$, and similarly for cross-interactions $A_{ij} = \mu$ ($i\neq j$).
Defining 
\begin{equation}
    \psi(x) \equiv (\mu - \mu_s) -  \left(\frac{f'(x)}{f(x)} - \frac{g'(x)}{g(x)}\right)\frac{f(x)}{g(x)h'(x)},
    \label{eq: psi}
\end{equation}
we find that stability of the homogeneous equilibrium $x_i^* = x^*$ requires $\psi(x^*) > 0$. This condition involves the relative strength of diagonal and off-diagonal interactions ($\mu_s - \mu$), but also on the relative growth rate of the production and response functions near the equilibrium ($f'(x_*)/f(x_*) - g'(x_*)/g(x_*)$).

With power laws, the condition $\psi(x^*)<0$ evaluates to 
\begin{equation}
    (\alpha - \beta)(N-1) < \gamma(\mu_s/\mu- 1) - (\alpha - \beta)(\mu_s/\mu),
\end{equation}
leading to three different regimes:
\begin{itemize}
    \item If $\alpha = \beta$, stability requires $\mu_s > \mu$, i.e.
    self-interactions must be stronger than cross-interactions.
    This is the usual conclusion drawn from the Lotka-Volterra model.
    \item If $\alpha > \beta$, stability places an upper bound on $N$: the more complex the system, the less likely to be stable.
    We can call this ``May" behavior.
    \item If $\alpha < \beta$, stability places an lower bound on $N$: the more complex the system, the more likely to be stable.
    This is ``anti-May" behavior.
\end{itemize}

>>>>

\emph{Homogeneous interactions.---}Under what condition does~\eqref{dynamics} admit a (linearly) stable equilibrium? 
We begin with the simple case where all self-interactions have the same strength $A_{ii} = \mu_s$, and similarly for cross-interactions $A_{ij} = \mu$ ($i\neq j$).
Defining 
\begin{equation}
    \psi(x) \equiv (\mu_s - \mu) -  \left(\frac{f'(x)}{f(x)} - \frac{g'(x)}{g(x)}\right)\frac{f(x)}{g(x)h'(x)},
    \label{eq: psi}
\end{equation}
we find that stability of the homogeneous equilibrium $x_i^* = x^*$ requires $\psi(x^*)g(x^*)h'(x^*) > 0$. This reduces to $\psi(x^*)>0$ if we assume (and we will, for simplicity, in the following) $g$ and $h$ are positive, increasing functions. This condition involves the relative strength of diagonal and off-diagonal interactions ($\mu_s - \mu$), but also on the relative growth rate of the production and response functions near the equilibrium ($f'(x^*)/f(x^*) - g'(x^*)/g(x^*)$).

With power laws, the condition $\psi(x^*)>0$ evaluates to 
\begin{equation}
    (\alpha - \beta)(N-1) < \gamma(\mu_s/\mu- 1) - (\alpha - \beta)(\mu_s/\mu),
\end{equation}
leading to three different regimes:
\begin{itemize}
    \item If $\alpha = \beta$, stability requires $\mu_s > \mu$, i.e.
    self-interactions must be stronger than cross-interactions.
    This is the usual conclusion drawn from the Lotka-Volterra model.
    \item If $\alpha > \beta$, stability places an upper bound on $N$: the more complex the system, the less likely to be stable.
    We can call this ``May" behavior.
    \item If $\alpha < \beta$, stability places an lower bound on $N$: the more complex the system, the more likely to be stable.
    This is ``anti-May" behavior.
\end{itemize}

### 2023-05-05 papers/model.tex
add link to repo
update draft

\emph{Model.---}Consider the dynamical system in $N$ variables
\begin{equation}\label{dynamics}
    \dot{x}_i = f(x_i) + \sum_{j}A_{ij}g(x_i)h(x_j) \, .
\end{equation}
Here $f(x_i)$ represents the self-interaction of a population $i$, while $g(x_i)$ and $h(x_j)$ capture the cross-interaction of $i$ with other populations. That interaction is weighted by a coefficient $A_{ij}$, such that $A_{ij} > 0$ (resp.
$A_{ji} < 0$) implies a positive (resp.
negative) effect of $j$ on the growth of $i$. 
We assume all of $f$, $g$ and $h$ are increasing functions, and refer to $f$ as the ``production function'' and $g$ as the ``response function''.

This rather general setting has been used to study universality in network dynamics \cite{Barzel2013} and to construct minimal models for complex dynamics \cite{Barzel2015}, with applications ranging from biological~\cite{Alon2006,Karlebach2008} 
to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.

The classic generalized Lotka-Volterra (GLV) competition model studied recently by Bunin and collaborators \cite{bunin2017ecological, biroli2018marginally} corresponds to $f$, $g$, $h$ all linear. 
However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$.
From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$.
Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$~\cite{Brown2004}, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
For reasons that are not currently understood, a similar pattern of growth appears to recur at the community level~\cite{Hatton2015,Hatton2023}.
In a different direction, it has been recently suggested that predator-prey interactions can be modelled with a square root law ($g(x) \sim h(x) \sim x^{1/2}$)~\cite{Barbier2021,Mazzarisi2023}.

>>>>

\emph{Model.---}Consider the dynamical system in $N$ variables
\begin{equation}\label{dynamics}
    \dot{x}_i = f(x_i) - \sum_{j}A_{ij}g(x_i)h(x_j) \, .
\end{equation}
Here $f(x_i)$ represents the self-interaction of a population $i$, while $g(x_i)$ and $h(x_j)$ capture the cross-interaction of $i$ with other populations. That interaction is weighted by a coefficient $A_{ij}$, such that $A_{ij} > 0$ (resp.
$A_{ji} < 0$) implies a negative (resp.
positive) effect of $j$ on the growth of $i$. Non-zero diagonal elements $A_{ii}$ accounts for additional self-interaction with the same shape as cross-ones. 
We refer to $f$ as the ``production function'' and $g$ as the ``response function''.

This rather general setting has been used to study universality in network dynamics \cite{Barzel2013} and to construct minimal models for complex dynamics \cite{Barzel2015}, with applications ranging from biological~\cite{Alon2006,Karlebach2008} 
to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.
The classic generalized Lotka-Volterra (GLV) competition model studied recently by Bunin and collaborators \cite{bunin2017ecological, biroli2018marginally} corresponds to $f$, $g$, $h$ all linear. 
However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$.
From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$.
Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$~\cite{Brown2004}, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
For reasons that are not currently understood, a similar pattern of growth appears to recur at the community level~\cite{Hatton2015,Hatton2023}.
In a different direction, it has been recently suggested that predator-prey interactions can be modelled with a square root law ($g(x) \sim h(x) \sim x^{1/2}$)~\cite{Barbier2021,Mazzarisi2023}.

### 2023-05-05 papers/stability.tex
update draft

We compute the Jacobian matrix at equilibrium
\begin{align}
    J_{ij}^* & = A_{ij}g(x_i^*)h'(x_j^*) \qquad \qquad \textrm{for} \ i\neq j \label{eq: jac off-diag}\\
    J_{ii}^* & = f'(x_i^*) - g'(x_i^*)f(x_i^*)/g(x_i^*) + A_{ii}g(x_i^*)h'(x_j^*) \ , \label{eq: jac diag}
\end{align}
where we used $\sum_{j}A_{ij}h(x_j^*)=-f(x_i^*)/g(x_i^*)$.
In order to investigate the spectral properties of $J^*$, 
we follow Stone~\cite{Stone2018} in using a recent generalization of the circular law in random matrix theory~\cite{Ahmadian2015}.

@@

To use this result, we decompose the interaction matrix as $A = \mu\mathbf{1} + (\mu_s-\mu)I + S$,
with $\mathbf{1}$ the matrix with all entries equal to $1$ and $S$ a random matrix as defined above.
Up to the rank-one perturbation $\mu\mathbf{1}$ which does not affect stability properties \cite{Stone2018}, we can write the Jacobian~\eqref{eq: jac off-diag} as $J = M + LSR$ with diagonal matrices
\begin{equation}
    M_i = g(x_i^*)h'(x_i^*)\psi(x_i^*),\quad L = g(x_i^*), \quad R = h'(x_i^*),
\end{equation}
where $\psi(x)$ is the function defined in~\eqref{eq: psi}.

\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{figs/cavity.pdf}
    \caption{The equilibrium distribution with power-law self- and cross-interactions is accurately reproduced by 
    Eq.~\eqref{eq: dist general} and its Gaussian approximation.
    For simulations (in green) we used $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=50$, $\mu=\mu_s=10^{-2}$, $\sigma=5\cdot 10^{-3}$.}
    \label{fig: cavity sol.}
\end{figure}

Stability of the equilibrium requires that $\mathcal{D}$ be entirely contained in the left half-plane (all eigenvalues have negative real part). For this, it is no longer sufficient that $\psi(x_i^*) > 0$ for each $i$: we must also have that $0\notin \mathcal{D}$, hence from~\eqref{eq: domain}:
\begin{equation}
    \sum_{i=1}^N \Big(\frac{\sigma}{\psi(x_i^*)}\Big)^{2}
    < 1.
    \label{eq: random-stability}
\end{equation}
In the GLV model, $\psi(x) = \mu-\mu_s$ (independently of $x$) and we recover the classical condition $\sigma\sqrt{N} < (\mu_s-\mu)$ \footnote{In fact, the same condition holds when $f$ and $g$ are power laws with the same exponent.}. In general, the dependence on equilibrium values $x_i^*$ (which, in turn, can depend on $N$) does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition~\eqref{eq: random-stability}.

>>>>

We compute the Jacobian matrix at equilibrium
\begin{align}
    J_{ij}^* & = - A_{ij}g(x_i^*)h'(x_j^*) \qquad \qquad \textrm{for} \ i\neq j \label{eq: jac off-diag}\\
    J_{ii}^* & = f'(x_i^*) - g'(x_i^*)f(x_i^*)/g(x_i^*) - A_{ii}g(x_i^*)h'(x_j^*) \ , \label{eq: jac diag}
\end{align}
where we used $\sum_{j}A_{ij}h(x_j^*)=f(x_i^*)/g(x_i^*)$.
In order to investigate the spectral properties of $J^*$, 
we follow Stone~\cite{Stone2018} in using a recent generalization of the circular law in random matrix theory~\cite{Ahmadian2015}.

@@

To use this result, we decompose the interaction matrix as $A = \mu\mathbf{1} + (\mu_s-\mu)I + S$,
with $\mathbf{1}$ the matrix with all entries equal to $1$ and $S$ a random matrix as defined above.
Up to the rank-one perturbation $\mu\mathbf{1}$ which does not affect stability properties \cite{Stone2018}, we can write the Jacobian~\eqref{eq: jac off-diag} as $J = M + LSR$ with diagonal matrices
\begin{equation}
    M_i = - g(x_i^*)h'(x_i^*)\psi(x_i^*),\quad L = g(x_i^*), \quad R = h'(x_i^*),
\end{equation}
where $\psi(x)$ is the function defined in~\eqref{eq: psi}. 
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{figs/cavity.pdf}
    \caption{The equilibrium distribution with power-law self- and cross-interactions is accurately reproduced by 
    Eq.~\eqref{eq: dist general} and its Gaussian approximation.
    For simulations (in green) we used $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=50$, $\mu=\mu_s=10^{-2}$, $\sigma=5\cdot 10^{-3}$.}
    \label{fig: cavity sol.}
\end{figure}
Stability of the equilibrium requires that $\mathcal{D}$ be entirely contained in the left half-plane (all eigenvalues have negative real part). For this, it is no longer sufficient that $\psi(x_i^*) > 0$ for each $i$: we must also have that $0\notin \mathcal{D}$, hence from~\eqref{eq: domain}:
\begin{equation}
    \sum_{i=1}^N \Big(\frac{\sigma}{\psi(x_i^*)}\Big)^{2}
    < 1.
    \label{eq: random-stability}
\end{equation}
In the GLV model, $\psi(x) = \mu_s - \mu$ (independently of $x$) and we recover the classical condition $\sigma\sqrt{N} < (\mu_s-\mu)$ \footnote{In fact, the same condition holds when $f$ and $g$ are power laws with the same exponent.}. In general, the dependence on equilibrium values $x_i^*$ (which, in turn, can depend on $N$) does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition~\eqref{eq: random-stability}.
